---
title: "Preliminary Analysis of Automated Facial Action Coding in Genuine and Performed Smiles"
subtitle: "Pilot data audit, variance structure, and a proposed analytic approach — circulated for feedback"
author:
  - name: "Delphine Dang"
    affiliation: "The University of Sydney"
    note: "With Dr Jie Kang · Circulated to Prof Ted Ruffman and Qi (Zoe) for comment"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-align: center
    fig-cap-location: bottom
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
---
```{=html}
<style>
/* ---- Report styling (single-file, no external SCSS) ---- */
@import url('https://fonts.googleapis.com/css2?family=Source+Serif+4:ital,opsz,wght@0,8..60,400;0,8..60,600;0,8..60,700&family=Source+Sans+3:wght@400;500;600;700&display=swap');

body { font-family: "Source Sans 3", -apple-system, sans-serif; color: #1F2937; line-height: 1.7; }
h1, h2, h3, h4 { font-family: "Source Serif 4", Georgia, serif; font-weight: 700; letter-spacing: -0.01em; color: #1E3A8A; }
h1:not(.title) { margin-top: 2.4rem; padding-bottom: 0.4rem; border-bottom: 2px solid #1E3A8A; }
h2 { margin-top: 1.8rem; color: #1E40AF; }
h3 { margin-top: 1.4rem; color: #2952c8; font-weight: 600; }
.header-section-number { color: #D97706; font-weight: 700; margin-right: 0.3rem; }

#title-block-header {
  margin-bottom: 2.5rem; padding: 2.4rem; border-radius: 14px;
  background: linear-gradient(135deg, #1E3A8A 0%, #1E40AF 55%, #312E81 100%);
  color: #EFF6FF;
}
#title-block-header .title { color: #fff; font-weight: 700; }
#title-block-header .subtitle, #title-block-header .quarto-title-meta, #title-block-header a { color: #DBEAFE; }

a { color: #1E40AF; text-decoration: none; border-bottom: 1px solid rgba(30,64,175,.25); }
a:hover { color: #D97706; border-bottom-color: #D97706; }

nav[role="doc-toc"] a { border-bottom: none; color: #475569; font-size: .9rem; }
nav[role="doc-toc"] .active { color: #1E3A8A !important; font-weight: 600; border-left: 2px solid #D97706; padding-left: .5rem; }

table { font-size: .9rem; border-collapse: collapse; margin: 1.5rem auto; }
thead th { background: #1E3A8A; color: #fff; font-weight: 600; padding: .5rem .75rem; }
tbody td { padding: .45rem .75rem; border-bottom: 1px solid #E5E7EB; }
tbody tr:nth-child(even) { background: #F8FAFC; }

figure img { border: 1px solid #E5E7EB; border-radius: 8px; box-shadow: 0 1px 3px rgba(0,0,0,.06), 0 4px 12px rgba(0,0,0,.04); }
figcaption { font-size: .85rem; color: #6B7280; text-align: center; margin-top: .6rem; font-style: italic; }

code { background: #F1F5F9; color: #BE123C; padding: .1rem .35rem; border-radius: 4px; }
blockquote { border-left: 3px solid #D97706; background: #F8FAFC; padding: .8rem 1.2rem; }
p strong:first-child { color: #1E3A8A; }
main { max-width: 860px; }


/* ---- Prominent Code-tools button (top-right "</> Code" menu) ---- */
#quarto-code-tools-menu, .code-tools-button, a.code-tools-button,
button.code-tools-button, #quarto-code-tools-source,
.quarto-title-tools, .quarto-code-tools-button {
  display: inline-flex !important; align-items: center;
  background: #D97706 !important; color: #ffffff !important;
  border: 2px solid #B45309 !important; border-radius: 8px !important;
  padding: 0.4rem 0.9rem !important; font-weight: 700 !important;
  font-family: "Source Sans 3", sans-serif !important; font-size: 0.95rem !important;
  text-decoration: none !important; box-shadow: 0 2px 6px rgba(217,119,6,.35) !important;
  transition: background .15s ease, transform .15s ease; z-index: 1000;
}
#quarto-code-tools-menu:hover, .code-tools-button:hover,
a.code-tools-button:hover, button.code-tools-button:hover,
#quarto-code-tools-source:hover, .quarto-code-tools-button:hover {
  background: #B45309 !important; color: #fff !important;
  transform: translateY(-1px); box-shadow: 0 4px 10px rgba(217,119,6,.45) !important;
}
.code-tools-button .bi, #quarto-code-tools-menu .bi { color: #fff !important; }

/* ---- Code-tools dropdown menu (Show All / Hide All Code / View Source) ---- */
/* The menu that opens on click: black text on a white background */
.quarto-code-tools-button + .dropdown-menu,
#quarto-code-tools-menu + .dropdown-menu,
.code-tools-button + .dropdown-menu,
ul.dropdown-menu[aria-labelledby*="code-tools"],
.dropdown-menu.show {
  background-color: #ffffff !important;
  border: 1px solid #CBD5E1 !important;
  border-radius: 8px !important;
  box-shadow: 0 4px 14px rgba(0,0,0,.12) !important;
  padding: 0.25rem 0 !important;
}
.quarto-code-tools-button + .dropdown-menu .dropdown-item,
#quarto-code-tools-menu + .dropdown-menu .dropdown-item,
.code-tools-button + .dropdown-menu .dropdown-item,
ul.dropdown-menu[aria-labelledby*="code-tools"] .dropdown-item,
.dropdown-menu.show .dropdown-item {
  color: #000000 !important;
  background-color: #ffffff !important;
  font-family: "Source Sans 3", sans-serif !important;
  font-weight: 500 !important;
  padding: 0.4rem 1rem !important;
}
/* Hover/focus state: keep text black, light grey background for feedback */
.quarto-code-tools-button + .dropdown-menu .dropdown-item:hover,
.quarto-code-tools-button + .dropdown-menu .dropdown-item:focus,
#quarto-code-tools-menu + .dropdown-menu .dropdown-item:hover,
#quarto-code-tools-menu + .dropdown-menu .dropdown-item:focus,
.code-tools-button + .dropdown-menu .dropdown-item:hover,
.code-tools-button + .dropdown-menu .dropdown-item:focus,
ul.dropdown-menu[aria-labelledby*="code-tools"] .dropdown-item:hover,
ul.dropdown-menu[aria-labelledby*="code-tools"] .dropdown-item:focus,
.dropdown-menu.show .dropdown-item:hover,
.dropdown-menu.show .dropdown-item:focus {
  color: #000000 !important;
  background-color: #F1F5F9 !important;
}
</style>
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center"
)
library(knitr)
library(lme4)

OUT <- "results_outputs"
man    <- read.csv(file.path(OUT, "01_file_manifest.csv"),     stringsAsFactors = FALSE)
clip   <- read.csv(file.path(OUT, "03_clip_summary_long.csv"), stringsAsFactors = FALSE)
adat   <- read.csv(file.path(OUT, "07_analysis_dataset.csv"),  stringsAsFactors = FALSE)
pow    <- read.csv(file.path(OUT, "12_power_simulation.csv"),  stringsAsFactors = FALSE)
frames <- read.csv(file.path(OUT, "02_frames_wide.csv"),       stringsAsFactors = FALSE)

ok  <- man[man$status == "ok", ]
gen <- adat[adat$smile_type == "Genuine", ]
n_files <- nrow(man); n_ok <- nrow(ok); n_state <- sum(man$status == "state_log_only")

au12 <- clip[clip$channel == "AU12", ]; au06 <- clip[clip$channel == "AU06", ]
cens_end_12   <- mean(au12$censored_end)
cens_start_12 <- mean(au12$censored_start)
cens_end_06   <- mean(au06$censored_end)

CONFIRM    <- c("AU06_pk", "AU12_pk", "duchenne_prop", "Happy_pk")
null_clips <- gen[is.na(gen$duchenne_prop), ]
n_null     <- nrow(null_clips)

NAVY <- "#1E3A8A"; AMBER <- "#D97706"; RED <- "#BE123C"; GREY <- "#CBD5E1"
```

::: {.callout-note appearance="simple" icon="false"}
**📋 Purpose** A preliminary analysis of the pilot FaceReader exports,
circulated for comment rather than approval. It reports what the data can
and cannot currently support, and proposes an analytic approach for the
full dataset. All figures and statistics are rebuilt from the extraction
pipeline at render time. The genuine-versus-performed contrast has not
been run, as those data have not yet arrived.
:::
::: {.callout-note appearance="simple"}
**Reproducible code.** This report renders from small committed summary files
in the folder `results_outputs`; the raw data is not needed. Every statistic, table and figure
below, including all plots, is rebuilt live from those summaries
when the document is rendered. Full code, data summaries:
<https://github.com/delphinedang/Smile_Project>. Session information is in the
Appendix.
:::

# Executive Summary {#sec-exec-summary}

Three findings from the 108 pilot files bear on how the full dataset
should be analysed.

**Repeated smiles from one person are strongly dependent.** The
between-smiler intraclass correlation on AU12 peak is 0.64, 95% CI
[0.37, 0.84]: individual
smilers differ by up to 3.6 grades on a 0–5 scale (@fig-caterpillar).
The 93 usable genuine smiles therefore carry the information of
approximately 23 independent observations. Linear mixed models are
required; repeated-measures ANOVA, which would demand person-level
averaging, is not appropriate for these data.

**Peak intensity is confounded with clip length.** Peak AU12 rises with
clip duration (*β* = 0.26 per SD, 95% CI [0.08, 0.44]), but this is an artefact of
maximising over a longer sample: truncating every clip to a common window
removes the association entirely (*p* = .76, @fig-duration). Unless clip
lengths are matched across conditions, peak measures require duration as
a covariate.

**Duration measures are heavily right-censored.** AU12 remains active in
the final frame of `r sprintf("%.0f%%", 100*cens_end_12)` of clips, so
recorded durations are lower bounds rather than measurements. Onset,
offset and speed are not recoverable from the current exports.

Two data-quality issues also require decisions: `r n_state` files were
exported in a format containing no action-unit data, and `r n_null`
clips labelled genuine contain no detectable smile.

::: {.callout-important appearance="simple" icon="false"}
**Points for discussion.** (i) The four candidate outcome measures in
@sec-outcomes, and whether `Happy_pk` should be replaced given it
correlates at *r* = 0.85 with `AU12_pk`. (ii) Whether clips can be re-cut
from the original recordings with padding, which would recover onset,
offset and duration. (iii) Treatment of the `r n_null` smile-less clips.
(iv) Whether the culture × smile-type interaction should be framed as
exploratory, given it reaches only 31% power at 30 smilers per culture
(@sec-power).
:::

# Data and Methods {#sec-methods}

## Materials {#sec-materials}

FaceReader detailed logs for 108 clips, each a manually segmented smile
episode from a longer recording. Each frame carries eight emotion
probabilities (0–1) and intensity grades for 20 action units
(NotActive / A–E, recoded 0–5). Sampling is 8 fps.

Clips are identified by an internal code (`E11HA01-01`) giving culture,
smiler, stimulus type (HA = humorous, SA = sad) and take. Filenames are
unreliable: the `Participant_###` fields are upload identifiers, and the
sequential clip number does not always match the stimulus.

## Design {#sec-design}

The unit of analysis is one smile episode. Smile type is within-smiler;
culture is between-smiler; stimulus videos recur across smilers. The
design is unbalanced — smilers contributed 3–6 clips each, and some
produced none of a given type.

## Statistical model {#sec-model}

$$
y_{ijk} = \beta_0 + \beta_1 \text{type}_{ijk} + \beta_2 \text{culture}_j
        + \beta_3 (\text{type} \times \text{culture})_{ijk}
        + u_j + v_k + \varepsilon_{ijk}
$$

with $u_j \sim N(0, \sigma^2_u)$ for smiler $j$, $v_k \sim N(0,
\sigma^2_v)$ for stimulus $k$, and $\varepsilon \sim N(0, \sigma^2)$.
Estimation by REML; inference on `type` and the interaction by likelihood
ratio test against the reduced model fitted by ML.

Mixed models are indicated for three reasons. The design is unbalanced,
so repeated-measures ANOVA would require averaging within person —
weighting a smiler with one clip equally with one contributing eight.
Smilers with no clips in a condition are retained rather than deleted
listwise. And stimulus variance, if unmodelled, inflates Type I error for
the effects of interest.

# Results {#sec-results}

## Variance structure {#sec-variance}

```{r tbl-icc}
#| tbl-cap: "Variance components from intercept-only mixed models, genuine clips (n = 93, 16 smilers, 7 stimuli). ICC is the proportion of total variance at that level."
vc1 <- function(y, grp) {
  f <- gen[!is.na(gen[[y]]), ]
  m <- lmer(as.formula(paste(y, "~ 1 + (1|", grp, ")")), data = f, REML = TRUE)
  v <- as.data.frame(VarCorr(m))
  list(m = m, sd_b = sqrt(v$vcov[1]), sd_e = sqrt(v$vcov[2]),
       icc = v$vcov[1] / sum(v$vcov))
}
# Profile likelihood CI on the smiler ICC, propagated from the SD bounds.
icc_ci <- function(m) {
  ci <- suppressMessages(confint(m, method = "profile", oldNames = FALSE))
  sb <- ci["sd_(Intercept)|smiler_id", ]; se <- ci["sigma", ]
  c(sb[1]^2 / (sb[1]^2 + se[2]^2), sb[2]^2 / (sb[2]^2 + se[1]^2))
}
outs <- c("AU12_pk", "AU06_pk", "duchenne_prop", "Happy_pk", "AU12_prop")
tab <- do.call(rbind, lapply(outs, function(y) {
  a <- vc1(y, "smiler_id"); b <- vc1(y, "stimulus"); ci <- icc_ci(a$m)
  data.frame(Outcome = y,
             `SD smiler`    = round(a$sd_b, 3),
             `SD residual`  = round(a$sd_e, 3),
             `ICC smiler`   = round(a$icc, 3),
             `95% CI`       = sprintf("[%.2f, %.2f]", ci[1], ci[2]),
             `ICC stimulus` = round(b$icc, 3),
             check.names = FALSE)
}))
kable(tab, row.names = FALSE)
```

Between-smiler ICCs range from 0.44 to 0.69; stimulus ICCs from 0.00 to
0.05. Smiler identity accounts for most of the variance in every
candidate outcome, while stimulus accounts for almost none.

```{r fig-caterpillar}
#| fig-cap: "Estimated random intercepts for AU12 peak by smiler, with 95% prediction intervals, from `AU12_pk ~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus)`. Smilers span 3.6 grades on a 0–5 scale; those whose intervals exclude zero differ reliably from the sample average."
#| fig-height: 5.4
m12 <- lmer(AU12_pk ~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus),
            data = gen, REML = TRUE)
re  <- ranef(m12, condVar = TRUE)$smiler_id
est <- re[, 1]; names(est) <- rownames(re)
se  <- sqrt(as.numeric(attr(re, "postVar")))
o   <- order(est); est <- est[o]; se <- se[o]
par(mar = c(4.4, 4.6, 1, 1.5))
plot(est, seq_along(est), pch = 19, col = NAVY, yaxt = "n", ylab = "",
     xlim = range(c(est - 2*se, est + 2*se)),
     xlab = "Deviation from average smiler (AU12 grades)")
segments(est - 2*se, seq_along(est), est + 2*se, seq_along(est), col = NAVY)
axis(2, at = seq_along(est), labels = names(est), las = 1, cex.axis = .75)
abline(v = 0, lty = 2, col = RED)
```

```{r fig-varcomp}
#| fig-cap: "Partition of variance for each candidate outcome. Between-smiler variance dominates throughout, which is why the smiler random intercept cannot be omitted."
#| fig-height: 3.4
vp <- sapply(outs, function(y) { s <- vc1(y, "smiler_id"); c(s[["icc"]], 1 - s[["icc"]]) })
par(mar = c(4, 7.5, 0.6, 9), xpd = NA)
barplot(vp * 100, horiz = TRUE, las = 1, col = c(NAVY, GREY), border = NA,
        xlab = "% of variance", names.arg = outs, cex.names = .8)
legend(112, 4, c("Between smilers", "Within smiler"), fill = c(NAVY, GREY),
       border = NA, bty = "n", cex = .8)
```

## Peak intensity is confounded with clip length {#sec-duration}

Because the design is unbalanced in clip duration, we tested whether peak
measures depend on how long the clip is.

```{r tbl-duration}
#| tbl-cap: "Fixed effect of clip duration (standardised) on each candidate outcome, from `y ~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus)` fitted to genuine clips."
dur_tab <- do.call(rbind, lapply(CONFIRM, function(y) {
  d <- gen[!is.na(gen[[y]]), ]
  m <- lmer(as.formula(paste(y, "~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus)")),
            data = d, REML = TRUE)
  m0 <- lmer(as.formula(paste(y, "~ 1 + (1|smiler_id) + (1|stimulus)")),
             data = d, REML = FALSE)
  m1 <- update(m, REML = FALSE)
  lr <- anova(m0, m1)
  cf <- summary(m)$coefficients
  ci <- confint(m, parm = "beta_", method = "Wald")[2, ]
  data.frame(Outcome = y, n = nrow(d),
             Beta = round(cf[2, 1], 3),
             `95% CI` = sprintf("[%.3f, %.3f]", ci[1], ci[2]),
             `chi-sq` = round(lr$Chisq[2], 2), df = lr$Df[2],
             p = format.pval(lr$`Pr(>Chisq)`[2], digits = 2, eps = .001),
             check.names = FALSE)
}))
kable(dur_tab, row.names = FALSE)
```

Peak measures rise with clip duration; the scale-free `duchenne_prop`
does not. This is consistent with a sampling artefact rather than a
substantive effect — the maximum of a longer series is stochastically
larger. @fig-duration confirms it: restricting the peak to a fixed window
common to all clips removes the association.

```{r fig-duration}
#| fig-cap: "Left: peak AU12 computed over the whole clip rises with duration. Right: peak computed over the first 21 frames (2.6 s, the shortest clip) shows no such rise. In a mixed model the fixed-window association is negligible (β = 0.03, p = .76), indicating the left-hand pattern is an artefact of maximising over more samples."
#| fig-height: 3.6
tp <- sapply(gen$clip_id, function(id) {
  d <- frames[frames$clip_id == id, ]; max(d$AU12[1:21], na.rm = TRUE)
})
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
plot(gen$clip_duration_s, jitter(gen$AU12_pk, .4), pch = 19,
     col = adjustcolor(NAVY, .55), xlab = "Clip duration (s)",
     ylab = "AU12 peak", main = "Whole clip")
abline(lm(AU12_pk ~ clip_duration_s, gen), col = AMBER, lwd = 2)
plot(gen$clip_duration_s, jitter(tp, .4), pch = 19,
     col = adjustcolor(NAVY, .55), xlab = "Clip duration (s)",
     ylab = "AU12 peak, first 21 frames", main = "Fixed window")
abline(lm(tp ~ gen$clip_duration_s), col = AMBER, lwd = 2)
```

::: {.callout-important appearance="simple" icon="false"}
**Consequence for the analysis.** If genuine and performed smiles differ
systematically in clip length, an uncorrected comparison of peak
intensities will be biased. Clip duration should enter as a covariate for
all peak outcomes — not, as initially proposed, only for cumulative ones
— with a fixed-window peak reported as a sensitivity check.
:::

## Model diagnostics {#sec-diagnostics}

```{r fig-diagnostics}
#| fig-cap: "Diagnostics for the AU12 peak model. Residuals are approximately normal with mild discreteness from the six-level ordinal scale, and show no systematic relationship with fitted values. The discreteness motivates the ordinal sensitivity check proposed in Appendix 8.4."
#| fig-height: 3.4
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
r <- resid(m12, scaled = TRUE)
qqnorm(r, main = "Normal Q-Q", pch = 19, col = adjustcolor(NAVY, .5))
qqline(r, col = AMBER, lwd = 2)
plot(fitted(m12), r, pch = 19, col = adjustcolor(NAVY, .5),
     xlab = "Fitted", ylab = "Scaled residual", main = "Residuals vs fitted")
abline(h = 0, col = AMBER, lwd = 2, lty = 2)
```

## Data quality {#sec-audit}

```{r audit-counts, include=FALSE}
n_conflict <- sum(man$label_conflict == 1, na.rm = TRUE)
n_fitfail  <- sum(ok$n_fitfail_cells, na.rm = TRUE)
fps        <- unique(ok$frame_rate)[1]
```

Of `r n_files` files, `r n_ok` parsed successfully. Five issues arise.

**Right-censoring.** AU12 is active in the first frame of
`r sprintf("%.0f%%", 100*cens_start_12)` of clips and the final frame of
`r sprintf("%.0f%%", 100*cens_end_12)`; AU06 in
`r sprintf("%.0f%%", 100*cens_end_06)`. Recorded durations are therefore
lower bounds. Rise rate is computable for only
`r sum(!is.na(gen$au12_rise_rate))` of `r nrow(gen)` clips. Re-cutting
from the original recordings with approximately two seconds of padding
would recover onset, offset and duration.

```{r fig-raster}
#| fig-cap: "Every usable clip's AU12 and AU06 time course, one row per clip, time normalised to clip length and shade giving intensity grade. Two audit findings are visible at once: bands reaching the right-hand edge are right-censored, the offset never observed; and the pale rows at the foot of the AU12 panel are clips in which the lip-corner puller never activates."
#| fig-height: 5.8
raster_matrix <- function(ch, ids, k = 50) {
  t(sapply(ids, function(id) {
    v <- frames[[ch]][frames$clip_id == id]
    approx(seq(0, 1, length.out = length(v)), v,
           xout = seq(0, 1, length.out = k), method = "constant", rule = 2)$y
  }))
}
ord <- gen$clip_id[order(gen$AU12_prop)]
pal <- colorRampPalette(c("#F8FAFC", "#BFDBFE", "#60A5FA", NAVY))(6)
layout(matrix(c(1, 2, 3, 3), nrow = 2, byrow = TRUE), heights = c(6, 1))
par(mar = c(4.2, 3.4, 2.4, 1))
for (ch in c("AU12", "AU06")) {
  m <- raster_matrix(ch, ord)
  image(x = seq(0, 1, length.out = ncol(m)), y = seq_len(nrow(m)), z = t(m),
        col = pal, zlim = c(0, 5), axes = FALSE, xlab = "Proportion through clip",
        ylab = "", main = paste0(ch, if (ch == "AU12") " - Lip Corner Puller"
                                     else " - Cheek Raiser"))
  axis(1); box(col = "grey70")
  mtext("Clips, least to most active", side = 2, line = 1, cex = .75)
}
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", legend = c("None", "A", "B", "C", "D", "E"), fill = pal,
       border = "grey70", bty = "n", cex = .9, horiz = TRUE,
       title = "AU intensity grade")
layout(1)
```

```{r fig-censoring}
#| fig-cap: "Percentage of clips in which each action unit remains active in the final frame. Above the dashed line, recorded durations are lower bounds. AU12 and AU06, on which the hypotheses depend, are the most affected."
#| fig-height: 5.6
aus   <- grep("^AU[0-9]+_cens1$", names(adat), value = TRUE)
rates <- sort(sapply(aus, function(a) mean(adat[[a]], na.rm = TRUE)))
par(mar = c(4.4, 6.5, 1, 2))
cols <- ifelse(names(rates) %in% c("AU12_cens1", "AU06_cens1"), AMBER, NAVY)
barplot(rates * 100, horiz = TRUE, las = 1, xlim = c(0, 100), border = NA,
        names.arg = sub("_cens1", "", names(rates)), cex.names = .72,
        xlab = "% of clips still active in final frame", col = cols)
abline(v = 50, lty = 2, col = RED)
```

**Sampling rate.** All exports report `r fps` fps rather than the 30 fps
expected, quantising timing at 125 ms.

**Export format.** `r n_state` files are state logs containing only
dominant-expression transitions, with no action-unit data (all of C15,
all of C31, one C24 clip). Re-export would recover two smilers.

**Label conflict.** `r n_conflict` clip (`E25SA07-03`) is coded SA but
its source video is `Happy 7 CN`; it is excluded pending clarification.
A further `r n_fitfail` cells are `FIT_FAILED` and recoded to `NA`.

**Clips without a smile.** `r n_null` genuine-labelled clips show no AU12
activation in any frame.

```{r tbl-null}
#| tbl-cap: "Genuine-labelled clips in which AU12 never activates. Peaks are on the 0–5 ordinal scale; Happy is a probability."
nc <- null_clips[, c("clip_id", "smiler_id", "stimulus", "clip_duration_s",
                     "AU12_pk", "AU06_pk", "Happy_pk")]
nc$Happy_pk <- round(nc$Happy_pk, 3)
names(nc) <- c("Clip", "Smiler", "Stimulus", "Duration (s)",
               "AU12 peak", "AU06 peak", "Happy peak")
kable(nc, row.names = FALSE)
```

```{r null-compare, include=FALSE}
good <- gen[!is.na(gen$duchenne_prop), ]
n_aus <- function(ids) {
  s <- clip[clip$clip_id %in% ids & clip$channel_type == "action_unit", ]
  mean(tapply(s$peak, s$clip_id, function(x) sum(x > 0, na.rm = TRUE)))
}
au_null <- n_aus(null_clips$clip_id); au_good <- n_aus(good$clip_id)
hp_null <- mean(null_clips$Happy_pk); hp_good <- mean(good$Happy_pk)
```

These are not tracking failures: all have zero `FIT_FAILED` cells and
other action units activate normally, but they average
`r sprintf("%.1f", au_null)` active units against
`r sprintf("%.1f", au_good)` elsewhere, and peak happiness
`r sprintf("%.2f", hp_null)` against `r sprintf("%.2f", hp_good)`.
Mis-segmentation is the most likely explanation. Retaining them would
depress the genuine mean and inflate within-group variance; a rule
excluding clips with no AU12 activation, applied blind to condition and
reported with a sensitivity analysis, is proposed for discussion.

## Candidate outcome measures {#sec-outcomes}

| Measure | Variable | Rationale |
|------------------|-----------------|-------------------------------------|
| Cheek raiser, peak | `AU06_pk` | Duchenne marker; held to resist voluntary production |
| Lip corner puller, peak | `AU12_pk` | The smile action; expected in both types |
| Duchenne co-activation | `duchenne_prop` | Proportion of AU12-active frames also showing AU06 |
| Happiness, peak | `Happy_pk` | FaceReader composite |

: Candidate confirmatory outcomes {#tbl-confirm}

`duchenne_prop` conditions on AU12 being active, capturing temporal
co-occurrence rather than independent peaks, and is scale-free. It is
also the least correlated with the others (@sec-a3) and the only one
unaffected by the duration artefact (@tbl-duration).

The four are strongly intercorrelated: `AU12_pk` and `Happy_pk` at
*r* = 0.85, unsurprising given that FaceReader's happiness classifier is
substantially driven by lip-corner movement. Substituting a temporally
informative measure for `Happy_pk` would increase the independent
information in the set.

Remaining action units would be screened as exploratory with
Benjamini–Hochberg correction at *q* = .05. Fixing this split in advance
matters: across 20 action units, 8 emotion channels and roughly 10
measures each, approximately 300 tests are available.

## Statistical power {#sec-power}

```{r tbl-power}
#| tbl-cap: "Simulated power at α = .05 for a moderate effect (d = 0.5), 500 replicates, using variance components from the pilot data. Full grid in Appendix 8.2."
pt <- pow[pow$effect_d == 0.5, c("smilers_per_culture", "power_main", "power_interaction")]
names(pt) <- c("Smilers per culture", "Power: main effect", "Power: interaction")
kable(pt, row.names = FALSE)
```

The main effect is well powered (94% at 20 smilers per culture). The
culture × smile-type interaction is not: 31% at 30 smilers per culture,
and 61% even at *d* = 0.8. The asymmetry is structural — smile type
operates within smiler, whereas the interaction asks whether that
within-smiler effect differs between groups, a between-smiler contrast
where the design is thinnest.

## Current coverage {#sec-status}

```{r tbl-status}
#| tbl-cap: "Pilot coverage. The genuine-versus-performed contrast is not yet testable."
lab <- ifelse(is.na(ok$smile_type) | ok$smile_type == "", "Unresolved", ok$smile_type)
st  <- as.data.frame.matrix(table(ok$culture, lab))
kable(cbind(Culture = rownames(st), st), row.names = FALSE)
```

The pipeline runs in pilot mode, producing derived variables,
descriptives and diagnostics while reporting inferential models as
blocked. It switches to full mode automatically when the remaining cells
arrive. Validation against a simulated complete dataset with a known
effect injected into AU06 recovered that effect (Hedges' *g* = 1.23) and
`duchenne_prop` (*g* = 1.47), while returning null results for `AU12_pk`
and `Happy_pk`, where no effect was injected.

# Discussion {#sec-discussion}

Three results shape the analysis of the full dataset. The dependence
among repeated smiles is strong enough that effective sample size is
governed by smiler count, not smile count — 93 clips behaving as
approximately 23 independent observations. Peak intensity is confounded
with clip duration through a sampling artefact, requiring either a
covariate or a fixed measurement window. And censoring of the kind
present here makes onset, offset and duration unrecoverable without
re-cutting.

The last point converges with Qi's independent assessment of the
material. Clip boundaries do not correspond to true smile boundaries;
duration and speed reflect the eliciting event and individual expressive
style alongside genuineness. Where that analysis proposes representing
each smile as a trajectory in AU space, robust to temporal stretching and
uncertain boundaries, the diagnosis is the same and the proposed remedy
is complementary to the approach here. @sec-a5 sets out how the two might
be reconciled, and what the present data can support.

A limitation deserves emphasis. With 93 genuine and one performed clip,
nothing here speaks to whether genuine and performed smiles differ. The
contribution is a characterisation of the measurement properties of these
data, and the consequences for how the eventual comparison should be
specified.

# Points for Discussion {#sec-asks}

1.  **Outcome measures.** Are the four in @tbl-confirm appropriate?
    Should `Happy_pk` be replaced given *r* = 0.85 with `AU12_pk`?
2.  **Re-cutting.** Do the original full-length recordings survive?
    Padding of \~2 s would recover onset, offset and duration.
3.  **Smile-less clips.** Is the proposed exclusion rule acceptable?
4.  **Sampling rate.** Was 8 fps intended, or introduced in conversion?
5.  **State-log files.** Can the 13 files be re-exported?
6.  **`E25SA07-03`.** Humorous or sad stimulus?
7.  **Stimulus sets.** Did NZ European smilers view the same videos as
    Chinese smilers? This determines whether stimulus is crossed with or
    nested within culture.
8.  **Sample size.** How many smilers per culture in total?

# Appendix {#sec-appendix}

## Sample composition {#sec-a1}

```{r tbl-composition}
#| tbl-cap: "Genuine clips per smiler and per stimulus."
sm <- as.data.frame(table(gen$smiler_id)); names(sm) <- c("Smiler", "Clips")
sm <- sm[sm$Clips > 0, ]
sti <- as.data.frame(table(gen$stimulus)); names(sti) <- c("Stimulus", "Clips")
n <- max(nrow(sm), nrow(sti))
pad <- function(d, n) { d[] <- lapply(d, as.character)
  if (nrow(d) < n) d[(nrow(d)+1):n, ] <- ""; d }
kable(cbind(pad(sm, n), pad(sti, n)), row.names = FALSE, align = "lrlr")
```

```{r tbl-durstat}
#| tbl-cap: "Clip duration (seconds), genuine clips."
kable(as.data.frame(t(round(c(summary(gen$clip_duration_s)), 2))), row.names = FALSE)
```

## Power simulation {#sec-a2}

```{r tbl-powerfull}
#| tbl-cap: "Simulated power. Monte Carlo SE is about 2 percentage points at 500 replicates, 4 at 150."
pf <- pow
names(pf) <- c("Smilers per culture", "Effect d", "Power: main",
               "Power: interaction", "Replicates")
kable(pf, row.names = FALSE)
```

Each replicate simulates two cultures × *k* smilers, each contributing
3–6 genuine and 3–6 performed clips. Smiler intercepts are drawn from
N(0, 1.04²) and residuals from N(0, 0.885²), both pilot estimates for
`AU06_pk`. Genuine clips in culture A are shifted by *d* residual SDs and
in culture B by *d*/2. Reproduced by `03_power_simulation.R`.

## Correlations among candidate outcomes {#sec-a3}

```{r tbl-cor}
#| tbl-cap: "Pearson correlations, genuine clips (n = 93)."
cm <- cor(gen[, CONFIRM], use = "pairwise.complete.obs")
kable(round(cm, 3))
```

Correlations run from `r sprintf("%.2f", min(cm[upper.tri(cm)]))` to
`r sprintf("%.2f", max(cm[upper.tri(cm)]))`. Correlated tests inflate
familywise error less than independent ones, so leaving the confirmatory
set uncorrected is defensible; but the set also carries less independent
information than four measures implies.

## Censoring by action unit {#sec-a4}

```{r tbl-censfull}
#| tbl-cap: "Start-censored = active in the first frame (onset unobserved); end-censored = active in the final frame (offset unobserved). All 95 usable clips."
aud <- clip[clip$channel_type == "action_unit", ]
ct <- do.call(rbind, lapply(split(aud, aud$channel), function(d) {
  data.frame(AU = d$channel[1],
             `Start n` = sum(d$censored_start),
             `Start %` = round(100 * mean(d$censored_start), 1),
             `End n` = sum(d$censored_end),
             `End %` = round(100 * mean(d$censored_end), 1),
             check.names = FALSE)
}))
kable(ct[order(-ct$`End %`), ], row.names = FALSE)
```

## Relation to trajectory-based approaches {#sec-a5}

Qi's proposal represents each smile as an oriented trajectory in
AU-intensity space, analysed under reparameterisation so that ordering
and coordination — rather than execution speed — carry the signal. The
motivating diagnosis matches the censoring and duration findings above.
Three properties of the present data bear on feasibility.

**Temporal resolution.** At 8 fps the median clip contains 40 frames.
Elastic functional methods and higher-order signature representations
generally assume denser sampling; with 40 points in a 20-dimensional
space the representation would be sparsely determined. Re-export at
30 fps would materially improve this.

**Measurement scale.** FaceReader detailed logs give ordinal grades
(NotActive, A–E), and grade E is almost unused, leaving five working
levels. Trajectories are therefore step functions rather than smooth
curves, which weakens methods assuming differentiable paths. OpenFace,
referenced in Qi's note, returns continuous AU intensities and would suit
trajectory methods considerably better. Confirming which software will
supply the final measurements would be useful, as the choice constrains
the method.

**Dimensionality against sample size.** With approximately 23 effective
independent observations, high-dimensional trajectory descriptors would
be estimated with substantial uncertainty.

A tractable intermediate exists. Restricting attention to the two or
three action units the hypothesis names — AU06, AU12, optionally AU25 —
makes a low-order path-signature representation feasible: for $d$
channels the level-2 signature has $d + d^2$ terms, so 12 features for
$d = 3$. The level-2 antisymmetric terms encode lead–lag ordering
directly, which is precisely the Duchenne question in temporal form: does
cheek raise precede, accompany, or follow lip-corner movement, and does
that ordering differ between genuine and performed smiles? Such features
are scale-free and reparameterisation-invariant, so they inherit the
robustness to boundary and speed variation that motivates the trajectory
framing, while entering the same mixed model as the outcomes above.

This narrower version is estimable with the present data and testable
within the existing design. The fuller multi-scale programme is better
suited to 30 fps continuous-intensity data on the complete sample, and
could be developed once those are available.

## Coding decisions {#sec-a6}

| Decision | Rule |
|-------------------|-----------------------------------------------------|
| AU intensity | NotActive = 0, A–E = 1–5, `FIT_FAILED` = `NA` |
| AU active | Intensity ≥ 1 |
| Emotion active | Probability ≥ 0.5 |
| Smile type | HA stimulus = genuine, SA = performed |
| Duration covariate | All peak, timing and cumulative outcomes (`_pk`, `_tpk`, `_durtot`, `_durmax`, `_auc`, `_neps`). Revised following @sec-duration; scale-free proportions exempt |
| Smiler and stimulus | From the internal clip code, not the export filename |
| Multiplicity | Confirmatory uncorrected; exploratory Benjamini–Hochberg at *q* = .05 |

: Coding decisions applied {#tbl-coding}

## Pipeline {#sec-a7}

Three scripts, base R with `lme4`. `01_extract_facereader.R` parses raw
exports (outputs 01–06); `02_analysis.R` derives measures and fits models
(07–11); `03_power_simulation.R` produces the power grid (12).
`04_clip_summary_wide.csv` is the SPSS import file, with variable names
of at most 16 characters. Full documentation in `README.md`.

## Session information {#sec-a8}

```{r session}
sessionInfo()
```
