---
title: "Preliminary analysis of automated facial action coding in genuine and performed smiles"
author: "Delphine Dang, supervised by Dr Jie Kang"
affiliation: "The University of Sydney"
date: today
page-layout: article
sidebar: false
number-sections: true
number-depth: 3
format:
  html:
    theme: united
    highlight-style: tango
    toc: true
    toc-title: "Table of Contents"
    toc-location: right
    toc-expand: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    title-block-banner: "#e64626"
    css: styles.css
execute:
  eval: true
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(knitr)
library(lme4)

OUT <- "results_outputs"
man    <- read.csv(file.path(OUT, "01_file_manifest.csv"),     stringsAsFactors = FALSE)
clip   <- read.csv(file.path(OUT, "03_clip_summary_long.csv"), stringsAsFactors = FALSE)
adat   <- read.csv(file.path(OUT, "07_analysis_dataset.csv"),  stringsAsFactors = FALSE)
pow    <- read.csv(file.path(OUT, "12_power_simulation.csv"),  stringsAsFactors = FALSE)
frames <- read.csv(file.path(OUT, "02_frames_wide.csv"),       stringsAsFactors = FALSE)

ok  <- man[man$status == "ok", ]
gen <- adat[adat$smile_type == "Genuine", ]
n_files <- nrow(man); n_ok <- nrow(ok); n_state <- sum(man$status == "state_log_only")

au12 <- clip[clip$channel == "AU12", ]; au06 <- clip[clip$channel == "AU06", ]
cens_end_12   <- mean(au12$censored_end)
cens_start_12 <- mean(au12$censored_start)
cens_end_06   <- mean(au06$censored_end)

CONFIRM    <- c("AU06_pk", "AU12_pk", "duchenne_prop", "Happy_pk")
null_clips <- gen[is.na(gen$duchenne_prop), ]
n_null     <- nrow(null_clips)
n_conflict <- sum(man$label_conflict == 1, na.rm = TRUE)
fps        <- unique(ok$frame_rate)[1]

# USYD palette, to sit with the banner
USYD <- "#E64626"; SLATE <- "#37474F"; LIGHT <- "#CFD8DC"; MID <- "#78909C"

# Mean number of active action units, smile-less clips vs the rest
n_active_aus <- function(ids) {
  s <- clip[clip$clip_id %in% ids & clip$channel_type == "action_unit", ]
  mean(tapply(s$peak, s$clip_id, function(x) sum(x > 0, na.rm = TRUE)))
}
au_null <- n_active_aus(null_clips$clip_id)
au_good <- n_active_aus(gen$clip_id[!is.na(gen$duchenne_prop)])
```

**This report was prepared for Prof Ted Ruffman (Department of Psychology,
University of Otago) by the CPC Data Science Hub, University of Sydney
(v0.1; last updated `r format(Sys.Date(), "%d %B %Y")`).**

::: {.callout-note appearance="simple"}
**Reproducible code.** This report renders from small committed summary files
in the folder `results_outputs`; the raw data is not needed. Every statistic, table and figure
below, including all plots, is rebuilt live from those summaries
when the document is rendered. Full code, data summaries:
<https://github.com/delphinedang/Smile_Project>. Session information is in the
Appendix.
:::

# Executive summary

- **Repeated smiles from one person are strongly dependent.** The
  between-smiler intraclass correlation on AU12 peak is 0.64 (95% CI
  [0.37, 0.84]). The 93 usable genuine clips carry the information of
  approximately 23 independent observations, so linear mixed models are
  required and repeated-measures ANOVA is not appropriate.
- **Peak intensity is confounded with clip length.** Peak AU12 rises
  with clip duration (*β* = 0.26, 95% CI [0.08, 0.44]), an artefact of
  maximising over a longer sample. Clip duration must be covaried, or
  peaks computed over a fixed window.
- **Onset, offset and duration are not recoverable.** AU12 remains
  active in the final frame of
  `r sprintf("%.0f%%", 100*cens_end_12)` of clips. Re-cutting from the
  original recordings would recover all three.
- **Two data-quality issues need decisions.** `r n_state` files were
  exported without action-unit data, and `r n_null` clips labelled
  genuine contain no detectable smile.
- **Recommended next step.** Confirm whether the original recordings can
  be re-cut, agree the treatment of the `r n_null` smile-less clips, and
  frame the culture × smile-type interaction as exploratory given it
  reaches only 31% power at 30 smilers per culture.

# Findings

## Key finding one: How much do repeated smiles from the same person resemble one another? {#sec-f1}

-   **Method / Approach / Strategy**

    Intercept-only linear mixed models were fitted to the 93 genuine
    clips, separately with smiler and with stimulus as the grouping
    factor, and the intraclass correlation computed from the resulting
    variance components. Confidence intervals are profile likelihood.

-   **Observation**

    Between-smiler ICCs range from 0.44 to 0.69 across candidate
    outcomes; stimulus ICCs from 0.00 to 0.05. Individual smilers differ
    by up to 3.6 grades on the 0–5 AU12 scale.

-   **Comments**

    Smiler identity accounts for most of the variance in every candidate
    outcome, while the particular video watched accounts for almost
    none. With an ICC of 0.64 and 5.8 clips per smiler the design effect
    is about 4.1, so effective sample size is governed by the number of
    smilers rather than the number of smiles.

-   **Recommendation**

    Retain the smiler random intercept in all models. Repeated-measures
    ANOVA, which would require averaging within person, should not be
    used. The stimulus random intercept may be dropped on evidence if
    SPSS fails to converge.

::: panel-tabset
### Main result

```{r fig-caterpillar}
#| fig-cap: "Random intercepts for AU12 peak by smiler with 95% prediction intervals. Smilers span 3.6 grades on a 0–5 scale; those whose interval excludes zero differ reliably from the sample average."
#| fig-height: 5
m12 <- lmer(AU12_pk ~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus),
            data = gen, REML = TRUE)
re  <- ranef(m12, condVar = TRUE)$smiler_id
est <- re[, 1]; names(est) <- rownames(re)
se  <- sqrt(as.numeric(attr(re, "postVar")))
o   <- order(est); est <- est[o]; se <- se[o]
par(mar = c(4.4, 4.6, 1, 1.5))
plot(est, seq_along(est), pch = 19, col = SLATE, yaxt = "n", ylab = "",
     xlim = range(c(est - 2*se, est + 2*se)),
     xlab = "Deviation from average smiler (AU12 grades)")
segments(est - 2*se, seq_along(est), est + 2*se, seq_along(est), col = SLATE)
axis(2, at = seq_along(est), labels = names(est), las = 1, cex.axis = .75)
abline(v = 0, lty = 2, col = USYD, lwd = 2)
```

### Supporting detail

```{r tbl-icc}
#| tbl-cap: "Variance components from intercept-only mixed models, genuine clips (n = 93, 16 smilers, 7 stimuli)."
vc1 <- function(y, grp) {
  f <- gen[!is.na(gen[[y]]), ]
  m <- lmer(as.formula(paste(y, "~ 1 + (1|", grp, ")")), data = f, REML = TRUE)
  v <- as.data.frame(VarCorr(m))
  list(m = m, sd_b = sqrt(v$vcov[1]), sd_e = sqrt(v$vcov[2]),
       icc = v$vcov[1] / sum(v$vcov))
}
icc_ci <- function(m) {
  ci <- suppressMessages(confint(m, method = "profile", oldNames = FALSE))
  sb <- ci["sd_(Intercept)|smiler_id", ]; se <- ci["sigma", ]
  c(sb[1]^2 / (sb[1]^2 + se[2]^2), sb[2]^2 / (sb[2]^2 + se[1]^2))
}
outs <- c("AU12_pk", "AU06_pk", "duchenne_prop", "Happy_pk", "AU12_prop")
kable(do.call(rbind, lapply(outs, function(y) {
  a <- vc1(y, "smiler_id"); b <- vc1(y, "stimulus"); ci <- icc_ci(a$m)
  data.frame(Outcome = y,
             `SD smiler`   = round(a$sd_b, 3),
             `SD residual` = round(a$sd_e, 3),
             `ICC smiler`  = round(a$icc, 3),
             `95% CI`      = sprintf("[%.2f, %.2f]", ci[1], ci[2]),
             `ICC stimulus`= round(b$icc, 3), check.names = FALSE)
})), row.names = FALSE)
```

```{r fig-varcomp}
#| fig-cap: "Partition of variance for each candidate outcome."
#| fig-height: 3
vp <- sapply(outs, function(y) { s <- vc1(y, "smiler_id"); c(s$icc, 1 - s$icc) })
par(mar = c(4, 7.5, 0.6, 9), xpd = NA)
barplot(vp * 100, horiz = TRUE, las = 1, col = c(SLATE, LIGHT), border = NA,
        xlab = "% of variance", names.arg = outs, cex.names = .8)
legend(112, 4, c("Between smilers", "Within smiler"), fill = c(SLATE, LIGHT),
       border = NA, bty = "n", cex = .8)
```
:::

## Key finding two: Does peak intensity measure the smile, or the length of the clip? {#sec-f2}

-   **Method / Approach / Strategy**

    Each candidate outcome was regressed on standardised clip duration
    in a mixed model with crossed random intercepts. To test whether any
    association was a sampling artefact, peak AU12 was recomputed over a
    fixed 21-frame window — the length of the shortest clip — and the
    model refitted.

-   **Observation**

    Peak measures rise with clip duration (AU12 *β* = 0.26, 95% CI
    [0.08, 0.44]; AU06 *β* = 0.28, [0.08, 0.47]). The scale-free
    `duchenne_prop` does not (*β* = 0.03, [−0.03, 0.09]). Over a fixed
    window the association disappears entirely (*β* = 0.03, *p* = .76).

-   **Comments**

    This is a property of the maximum, not of the smile: the largest
    value in a longer series is stochastically greater. If genuine and
    performed clips differ systematically in length, an uncorrected
    comparison of peak intensities will be biased in favour of whichever
    condition happens to have longer clips.

-   **Recommendation**

    Covary clip duration for all peak and timing outcomes, not only
    cumulative ones. Report a fixed-window peak as a sensitivity check.
    Prefer `duchenne_prop`, which is immune to the artefact.

::: panel-tabset
### Main result

```{r fig-duration}
#| fig-cap: "Left: peak AU12 over the whole clip rises with duration. Right: peak over the first 21 frames shows no such rise, indicating the left-hand pattern is an artefact of maximising over more samples."
#| fig-height: 3.6
tp <- sapply(gen$clip_id, function(id) {
  d <- frames[frames$clip_id == id, ]; max(d$AU12[1:21], na.rm = TRUE)
})
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
plot(gen$clip_duration_s, jitter(gen$AU12_pk, .4), pch = 19,
     col = adjustcolor(SLATE, .5), xlab = "Clip duration (s)",
     ylab = "AU12 peak", main = "Whole clip")
abline(lm(AU12_pk ~ clip_duration_s, gen), col = USYD, lwd = 2.4)
plot(gen$clip_duration_s, jitter(tp, .4), pch = 19,
     col = adjustcolor(SLATE, .5), xlab = "Clip duration (s)",
     ylab = "AU12 peak, first 21 frames", main = "Fixed window")
abline(lm(tp ~ gen$clip_duration_s), col = USYD, lwd = 2.4)
```

### Supporting detail

```{r tbl-duration}
#| tbl-cap: "Fixed effect of standardised clip duration on each candidate outcome. Tests are likelihood ratio against the reduced model."
kable(do.call(rbind, lapply(CONFIRM, function(y) {
  d  <- gen[!is.na(gen[[y]]), ]
  m  <- lmer(as.formula(paste(y, "~ scale(clip_duration_s) + (1|smiler_id) + (1|stimulus)")),
             data = d, REML = TRUE)
  m0 <- lmer(as.formula(paste(y, "~ 1 + (1|smiler_id) + (1|stimulus)")),
             data = d, REML = FALSE)
  lr <- anova(m0, update(m, REML = FALSE))
  ci <- confint(m, parm = "beta_", method = "Wald")[2, ]
  data.frame(Outcome = y, n = nrow(d),
             Beta = round(summary(m)$coefficients[2, 1], 3),
             `95% CI` = sprintf("[%.3f, %.3f]", ci[1], ci[2]),
             `chi-sq` = round(lr$Chisq[2], 2), df = lr$Df[2],
             p = format.pval(lr$`Pr(>Chisq)`[2], digits = 2, eps = .001),
             check.names = FALSE)
})), row.names = FALSE)
```

```{r fig-diagnostics}
#| fig-cap: "Diagnostics for the AU12 peak model. Residuals are approximately normal with mild discreteness from the ordinal scale, and unrelated to fitted values."
#| fig-height: 3.2
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))
r <- resid(m12, scaled = TRUE)
qqnorm(r, main = "Normal Q-Q", pch = 19, col = adjustcolor(SLATE, .5))
qqline(r, col = USYD, lwd = 2)
plot(fitted(m12), r, pch = 19, col = adjustcolor(SLATE, .5),
     xlab = "Fitted", ylab = "Scaled residual", main = "Residuals vs fitted")
abline(h = 0, col = USYD, lwd = 2, lty = 2)
```
:::

## Key finding three: Can onset, offset and duration be recovered from these clips? {#sec-f3}

-   **Method / Approach / Strategy**

    For every action unit, each clip was flagged as start-censored if
    the unit was active in the first frame, and end-censored if still
    active in the final frame, giving the proportion of clips for which
    onset or offset is unobserved.

-   **Observation**

    AU12 is start-censored in
    `r sprintf("%.0f%%", 100*cens_start_12)` of clips and end-censored
    in `r sprintf("%.0f%%", 100*cens_end_12)`; AU06 in
    `r sprintf("%.0f%%", 100*cens_end_06)`. Smile rise rate is
    computable for only `r sum(!is.na(gen$au12_rise_rate))` of
    `r nrow(gen)` clips. Sampling is `r fps` fps rather than the 30 fps
    expected, quantising timing at 125 ms.

-   **Comments**

    Clips were cut tightly around each smile, so recording typically
    begins after onset and ends before offset. Recorded durations are
    lower bounds rather than measurements. If censoring rates differ
    between conditions the comparison is misleading rather than merely
    imprecise.

-   **Recommendation**

    Re-cut from the original recordings with approximately two seconds
    of padding, which would recover onset, offset and duration. Failing
    that, use `prop_active` as the duration-like outcome and treat
    duration with survival methods. Confirm whether the 8 fps
    downsampling was intended.

::: panel-tabset
### Main result

```{r fig-censoring}
#| fig-cap: "Percentage of clips in which each action unit remains active in the final frame. AU12 and AU06, on which the hypotheses depend, are the most affected."
#| fig-height: 5.4
aus   <- grep("^AU[0-9]+_cens1$", names(adat), value = TRUE)
rates <- sort(sapply(aus, function(a) mean(adat[[a]], na.rm = TRUE)))
par(mar = c(4.4, 6.5, 1, 2))
cols <- ifelse(names(rates) %in% c("AU12_cens1", "AU06_cens1"), USYD, SLATE)
barplot(rates * 100, horiz = TRUE, las = 1, xlim = c(0, 100), border = NA,
        names.arg = sub("_cens1", "", names(rates)), cex.names = .72,
        xlab = "% of clips still active in final frame", col = cols)
abline(v = 50, lty = 2, col = MID)
```

### Supporting detail

```{r fig-raster}
#| fig-cap: "Every genuine clip's AU12 and AU06 time course, one row per clip, time normalised and shade giving intensity grade. Bands reaching the right-hand edge are right-censored; pale rows at the foot of the AU12 panel are clips with no lip-corner activation at all."
#| fig-height: 5.4
raster_matrix <- function(ch, ids, k = 50) {
  t(sapply(ids, function(id) {
    v <- frames[[ch]][frames$clip_id == id]
    approx(seq(0, 1, length.out = length(v)), v,
           xout = seq(0, 1, length.out = k), method = "constant", rule = 2)$y
  }))
}
ord <- gen$clip_id[order(gen$AU12_prop)]
pal <- colorRampPalette(c("#FFFFFF", "#FBD9CF", "#F4A28A", "#E64626", "#9E2A14"))(6)
layout(matrix(c(1, 2, 3, 3), nrow = 2, byrow = TRUE), heights = c(6, 1))
par(mar = c(4.2, 3.4, 2.4, 1))
for (ch in c("AU12", "AU06")) {
  m <- raster_matrix(ch, ord)
  image(x = seq(0, 1, length.out = ncol(m)), y = seq_len(nrow(m)), z = t(m),
        col = pal, zlim = c(0, 5), axes = FALSE,
        xlab = "Proportion through clip", ylab = "",
        main = paste0(ch, if (ch == "AU12") " - Lip Corner Puller"
                          else " - Cheek Raiser"))
  axis(1); box(col = "grey70")
  mtext("Clips, least to most active", side = 2, line = 1, cex = .75)
}
par(mar = c(0, 0, 0, 0)); plot.new()
legend("center", legend = c("None", "A", "B", "C", "D", "E"), fill = pal,
       border = "grey70", bty = "n", cex = .9, horiz = TRUE,
       title = "AU intensity grade")
layout(1)
```

```{r tbl-censfull}
#| tbl-cap: "Censoring by action unit across all 95 usable clips."
aud <- clip[clip$channel_type == "action_unit", ]
ct <- do.call(rbind, lapply(split(aud, aud$channel), function(d) {
  data.frame(AU = d$channel[1],
             `Start n` = sum(d$censored_start),
             `Start %` = round(100 * mean(d$censored_start), 1),
             `End n`   = sum(d$censored_end),
             `End %`   = round(100 * mean(d$censored_end), 1),
             check.names = FALSE)
}))
kable(ct[order(-ct$`End %`), ], row.names = FALSE)
```
:::

## Key finding four: Are all the clips usable? {#sec-f4}

-   **Method / Approach / Strategy**

    All `r n_files` exports were parsed and audited. Each clip records
    its stimulus twice — once as the source video name, once in the
    internal clip code — and the two were cross-checked. Clips were also
    screened for activation of AU12, the defining smile action.

-   **Observation**

    `r n_ok` files parsed. `r n_state` are state logs containing only
    dominant-expression transitions, with no action-unit data (all of
    C15, all of C31, one C24 clip). `r n_conflict` clip
    (`E25SA07-03`) is coded SA while its source video is `Happy 7 CN`.
    `r n_null` genuine-labelled clips show no AU12 activation in any
    frame.

-   **Comments**

    The smile-less clips are not tracking failures: all have zero
    `FIT_FAILED` cells and other action units activate normally, but
    they average `r sprintf("%.1f", au_null)`
    active units against `r sprintf("%.1f", au_good)` elsewhere, and peak happiness
    `r sprintf("%.2f", mean(null_clips$Happy_pk))` against
    `r sprintf("%.2f", mean(gen$Happy_pk[!is.na(gen$duchenne_prop)]))`.
    Mis-segmentation is the likely explanation. Retained, they would
    depress the genuine mean and inflate within-group variance.

-   **Recommendation**

    Re-export the `r n_state` state logs, which would recover two
    smilers. Clarify `E25SA07-03`. Adopt a rule excluding clips with no
    AU12 activation, applied blind to condition and reported with a
    sensitivity analysis both ways.

::: panel-tabset
### Main result

```{r tbl-null}
#| tbl-cap: "Genuine-labelled clips in which AU12 never activates. Peaks are on the 0–5 ordinal scale; Happy is a probability."
nc <- null_clips[, c("clip_id", "smiler_id", "stimulus", "clip_duration_s",
                     "AU12_pk", "AU06_pk", "Happy_pk")]
nc$Happy_pk <- round(nc$Happy_pk, 3)
names(nc) <- c("Clip", "Smiler", "Stimulus", "Duration (s)",
               "AU12 peak", "AU06 peak", "Happy peak")
kable(nc, row.names = FALSE)
```

### Supporting detail

```{r tbl-status}
#| tbl-cap: "Current design coverage. The genuine-versus-performed contrast is not yet testable."
lab <- ifelse(is.na(ok$smile_type) | ok$smile_type == "", "Unresolved", ok$smile_type)
st  <- as.data.frame.matrix(table(ok$culture, lab))
kable(cbind(Culture = rownames(st), st), row.names = FALSE)
```

```{r tbl-composition}
#| tbl-cap: "Genuine clips per smiler and per stimulus. The unbalance is why repeated-measures ANOVA is unsuitable."
sm <- as.data.frame(table(gen$smiler_id)); names(sm) <- c("Smiler", "Clips")
sm <- sm[sm$Clips > 0, ]
sti <- as.data.frame(table(gen$stimulus)); names(sti) <- c("Stimulus", "Clips")
n <- max(nrow(sm), nrow(sti))
pad <- function(d, n) { d[] <- lapply(d, as.character)
  if (nrow(d) < n) d[(nrow(d)+1):n, ] <- ""; d }
kable(cbind(pad(sm, n), pad(sti, n)), row.names = FALSE, align = "lrlr")
```
:::

## Key finding five: What will the study be able to detect? {#sec-f5}

-   **Method / Approach / Strategy**

    Power was estimated by simulation using variance components from the
    pilot data (between-smiler SD 1.04, residual SD 0.885), with 3–6
    clips per smiler per condition and the interaction set at half the
    main effect. 500 replicates per cell at *d* = 0.5.

-   **Observation**

    At a moderate effect the genuine-versus-performed main effect
    reaches 94% power with 20 smilers per culture and 99% with 30. The
    culture × smile-type interaction reaches 31% at 30 smilers per
    culture, and 61% even at *d* = 0.8.

-   **Comments**

    The asymmetry is structural. Smile type operates within smiler and
    is tested efficiently; the interaction asks whether that
    within-smiler effect differs between groups, a between-smiler
    contrast where the design is thinnest. A null interaction would
    therefore be ambiguous between a genuinely absent effect and one
    that could not be detected.

-   **Recommendation**

    Frame the interaction as exploratory, reporting effect size with
    confidence intervals rather than relying on significance, unless
    substantially more smilers per culture are available. Confirming the
    total number of smilers would settle this.

::: panel-tabset
### Main result

```{r tbl-power}
#| tbl-cap: "Simulated power at α = .05 for a moderate effect (d = 0.5), 500 replicates."
pt <- pow[pow$effect_d == 0.5, c("smilers_per_culture", "power_main", "power_interaction")]
names(pt) <- c("Smilers per culture", "Power: main effect", "Power: interaction")
kable(pt, row.names = FALSE)
```

### Supporting detail

```{r tbl-powerfull}
#| tbl-cap: "Full simulated power grid. Monte Carlo SE is about 2 percentage points at 500 replicates and 4 at 150, so small non-monotonicities are simulation noise."
pf <- pow
names(pf) <- c("Smilers per culture", "Effect d", "Power: main",
               "Power: interaction", "Replicates")
kable(pf, row.names = FALSE)
```

```{r tbl-cor}
#| tbl-cap: "Correlations among candidate outcomes, genuine clips (n = 93). AU12 peak and Happy peak are near-redundant."
cm <- cor(gen[, CONFIRM], use = "pairwise.complete.obs")
kable(round(cm, 3))
```

`AU12_pk` and `Happy_pk` correlate at
`r sprintf("%.2f", cm["AU12_pk","Happy_pk"])`, unsurprising given that
FaceReader's happiness classifier is substantially driven by lip-corner
movement. Substituting a temporally informative measure for `Happy_pk`
would increase the independent information in the confirmatory set.
:::

## Next steps

Ordered by who is blocked on what.

| Step | Action | Owner |
|------|-------------------------------------------|------------|
| 1 | Confirm whether original recordings can be re-cut with padding | Zoey |
| 2 | Re-export the `r n_state` state-log files as detailed logs | Zoey |
| 3 | Clarify the stimulus for `E25SA07-03` | Zoey |
| 4 | Confirm total smilers per culture, and whether NZ European smilers viewed the same videos | Zoey |
| 5 | Agree the exclusion rule for smile-less clips and the confirmatory outcome set | All |
| 6 | Transfer remaining three data cells via shared OneDrive | Zoey |
| 7 | Re-run pipeline in full mode; cross-check headline results in SPSS | Delphine |
| 8 | Circulate results before interpretation is written | All |

# Appendix

## Project overview

The study compares smiles produced under two conditions. Participants
watched humorous videos, producing spontaneous smiles, and sad videos,
during which they were instructed to smile regardless. Smile episodes
were segmented manually and processed through FaceReader. Smilers were
drawn from Chinese and New Zealand European backgrounds.

The analytic question is whether the two smile types separate on
objective facial measures, and whether that separation differs by the
smiler's cultural background. The second question is the theoretically
motivated one: if masking felt emotion is more culturally normative,
performed smiles from that group may be physically closer to genuine
ones, which would help explain existing emotion-recognition findings
rather than merely restating them.

A third question requires no new data. Human judgement data already
exists for these clips, so FaceReader measures can be related to
per-clip human accuracy — asking whether observers use the signal that
is objectively present.

## Data overview

`r n_files` FaceReader exports were received, of which `r n_ok` are
usable. Each is one manually segmented smile episode. Every frame
carries eight emotion probabilities (0–1) and intensity grades for 20
action units (NotActive, A–E, recoded 0–5). Sampling is `r fps` fps and
the median clip runs 5.0 s.

Current coverage is `r nrow(gen)` genuine clips and one performed clip
from 16 smilers, all Chinese. The remaining three cells — Chinese
performed, NZ European genuine, NZ European performed — are outstanding,
so the central contrast is not yet testable. The pipeline runs in pilot
mode, producing derived variables, descriptives and diagnostics while
reporting inferential models as blocked, and switches to full mode
automatically when those data arrive.

Clips are identified by an internal code (`E11HA01-01`) giving culture,
smiler, stimulus type and take. Export filenames are unreliable: the
`Participant_###` fields are upload identifiers rather than people, and
the sequential clip number does not always match the stimulus.

## Analytical rationale

The unit of analysis is one smile episode. Smile type is within-smiler,
culture between-smiler, and stimulus videos recur across smilers. The
proposed model is

$$
y_{ijk} = \beta_0 + \beta_1 \text{type}_{ijk} + \beta_2 \text{culture}_j
        + \beta_3 (\text{type} \times \text{culture})_{ijk}
        + u_j + v_k + \varepsilon_{ijk}
$$

with $u_j$ for smiler $j$, $v_k$ for stimulus $k$. Estimation by REML;
inference by likelihood ratio test against the reduced model fitted by
ML. Clip duration is covaried for peak, timing and cumulative outcomes.

Mixed models are indicated for three reasons. The design is unbalanced,
so repeated-measures ANOVA would require averaging within person,
weighting a smiler contributing one clip equally with one contributing
eight. Smilers with no clips in a condition are retained rather than
deleted listwise. And stimulus variance, if unmodelled, inflates Type I
error.

Four confirmatory outcomes are proposed — `AU06_pk`, `AU12_pk`,
`duchenne_prop` and `Happy_pk` — reported without multiplicity
correction as a small pre-specified set, with all remaining action units
screened as exploratory under Benjamini–Hochberg at *q* = .05. Fixing
this split before the data are seen matters: roughly 300 tests are
available across 20 action units, 8 emotion channels and 10 measures
each.

Action-unit grades are treated as continuous. This is an approximation —
the scale is ordinal and grade E is almost unused, leaving five working
levels — so the confirmatory outcomes should be re-run as ordinal mixed
models (cumulative link) as a sensitivity check.

## Relation to trajectory-based approaches

An alternative framing represents each smile as an oriented trajectory
in AU-intensity space, analysed under reparameterisation so that
ordering and coordination rather than execution speed carry the signal.
The motivating diagnosis matches the censoring and duration findings
above: clip boundaries do not correspond to smile boundaries, and speed
reflects the eliciting event and individual style alongside genuineness.

Three properties of the present data bear on feasibility. At `r fps` fps
the median clip holds 40 frames, where elastic functional methods and
higher-order signature representations generally assume denser sampling.
FaceReader detailed logs are ordinal with five working levels, so
trajectories are step functions rather than differentiable paths;
OpenFace returns continuous intensities and would suit these methods
considerably better, so confirming which software supplies the final
measurements would be useful. And with approximately 23 effective
independent observations, high-dimensional descriptors would carry
substantial uncertainty.

A tractable intermediate exists. Restricting attention to the action
units the hypothesis names — AU06, AU12, optionally AU25 — makes a
low-order path signature feasible: for $d$ channels the level-2
signature has $d + d^2$ terms, so 12 features for $d = 3$. The level-2
antisymmetric terms encode lead–lag ordering directly, which is the
Duchenne question in temporal form: does cheek raise precede, accompany
or follow lip-corner movement, and does that ordering differ by
condition? Such features are scale-free and
reparameterisation-invariant, inheriting the robustness that motivates
the trajectory framing, while entering the same mixed model as the
outcomes above.

## Glossary

```{r glossary}
glossary <- data.frame(rbind(

  c("Action Unit (AU)",
  "A single facial muscle movement in the Facial Action Coding System. FaceReader scores 20 of them per frame. AU06 (cheek raiser) and AU12 (lip corner puller) are central here."),

  c("Duchenne smile",
  "A smile involving both lip-corner movement (AU12) and cheek raise (AU06). The cheek raise is held to be difficult to produce voluntarily, making it a candidate marker of genuineness."),

  c("duchenne_prop",
  "Derived measure: of the frames in which AU12 is active, the proportion in which AU06 is also active. Captures co-occurrence rather than independent peaks, and is unaffected by clip length."),

  c("Intraclass correlation (ICC)",
  "The proportion of total variance attributable to a grouping factor. An ICC of 0.64 for smiler means 64% of variation lies between people rather than between their individual smiles."),

  c("Design effect",
  "The factor by which clustering inflates variance relative to independent sampling. A design effect of 4.1 means 93 clustered clips carry the information of about 23 independent observations."),

  c("Linear mixed model",
  "A regression model containing both fixed effects (smile type, culture) and random effects (smiler, stimulus), allowing correlated observations from the same person to be modelled rather than averaged away."),

  c("Random intercept",
  "A per-group deviation from the overall average, here allowing each smiler their own baseline expressiveness."),

  c("REML",
  "Restricted maximum likelihood, the default estimator for variance components in mixed models. Models differing in fixed effects must be compared under ML, not REML."),

  c("Right-censoring",
  "An event whose endpoint is not observed. Here, an action unit still active in the final frame has an unknown offset, so its recorded duration is a lower bound."),

  c("Benjamini-Hochberg FDR",
  "A multiplicity correction controlling the expected proportion of false positives among significant results. Less conservative than Bonferroni and appropriate for exploratory screens."),

  c("Confirmatory vs exploratory",
  "Confirmatory outcomes are specified before the data are seen and reported without correction; exploratory analyses are corrected and reported as hypothesis-generating."),

  c("Hedges' g",
  "A standardised mean difference with a small-sample correction. Used here to express effect magnitude alongside p-values."),

  c("State log vs detailed log",
  "Two FaceReader export formats. Detailed logs give per-frame emotion and action-unit values; state logs give only dominant-expression transitions and cannot be used for this analysis."),

  c("FIT_FAILED",
  "A FaceReader cell indicating the model could not fit that frame. Recoded to missing, never to zero."),

  c("Pilot mode / full mode",
  "The analysis script detects whether design coverage supports the planned models. In pilot mode it reports inferential models as blocked; it switches to full mode automatically when the remaining data arrive.")
), stringsAsFactors = FALSE)
names(glossary) <- c("Term", "Definition")

if (requireNamespace("DT", quietly = TRUE)) {
  DT::datatable(glossary, rownames = FALSE,
                options = list(dom = "Bfrtip", pageLength = 10, scrollX = TRUE))
} else {
  knitr::kable(glossary)
}
```

## R session info

```{r session}
sessionInfo()
```
